---
title: "Cost of Living in Toronto"
subtitle: "Are low-income families able to survive financially here?"
author: 
  - Hyunje Park
thanks: "Code and data are available at: https://github.com/davidpxrk/cost-of-living-toronto"
date: today
date-format: long
abstract: "This paper analyzes the living-cost data from low-income families living in Toronto, which is hosted publicly by the Open Data Toronto. It uses information on household income, expenses, government fiscal transfers and benefits to determine if low-income families are able to survive financially in Toronto. It was discovered that expenses exceeded income for most families (meaning low-income families were not able to make ends meet in Toronto), and rent, food costs were among the highest expenses. This finding is important as it shows cost disparity in expenses, and provide insights to fiscal policymakers on how to better support low-income families with their living costs."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
#install.packages('opendatatoronto')
#install.packages("dplyr")
#install.packages('tidyverse')
#install.packages('sf')
#remove.packages('knitr')
install.packages('xfun')
install.packages('knitr')


# RUN THIS IN THE TERMINAL IF THERE'S AN ERROR IN EXPORTING TO PDF
# Source: https://stackoverflow.com/questions/75906430/why-cant-i-render-any-quarto-documents-into-pdfs
# quarto install tool tinytex

library(here)
library(knitr)
library(opendatatoronto)
library(tidyverse)
library(dplyr)
library(sf)

## Read the 2 cleaned CSV files ##
analysis_data = read_csv(here::here('data/cleaned_data/cleaned_data.csv'))
  
analysis_data_inverted = read_csv(here::here('data/cleaned_data/cleaned_data_inverted.csv'))
```


# Introduction

Toronto is the biggest and the most populous city in Canada. As the fourth largest city in North America, Toronto serves as a catalyst when it comes to being an international hub for many businesses, arts, sports and diverse cultures, attracting people from all around the world to come to Toronto to study at some of the best institutions, work or even simply travel as a Tourist. More importantly, Toronto is the most diverse city in Canada, with many people of different cultures, backgrounds, ethnicity coexist, and bridge cultural gaps and create a unity. 

While this may seem like a perfect place to live for many international talents, residents of Toronto have said this to be far from the truth, with many sturggling financially to meet ends. 

# Data
```{r}

analysis_data
#kable(analysis_data)
#knitr::kable(head(analysis_data[, 1:4]), format = "pipe")
analysis_data %>% head() %>% knitr::kable()
```
## Data Source



## Overview of the Dataset


## Average Income of the 11 Households


## Average Expenses of the 11 Households


## Difference between Income and Expense

# Results

# Discussion

# Conclusion

# Appendix

# Statement on LLMs 

LLMs and other generative AI tools were not used in the making of this paper.

# References


